---
title: "initial_analysis"
author: "steph"
date: "`r Sys.Date()`"
output: html_document
---

# My first step is to find the question I want to research. 
# 1.I look through the data structure and the data form.
```{r}
read.csv("data/MonatszahlenBevölkerung.csv")
data<-read.csv("data/MonatszahlenBevölkerung.csv")
table(data$MONATSZAHL)
table(data$AUSPRAEGUNG)
```
# 2.List the broad categories of variables.
Nationality / Continents
Migration Background
Age Groups
Gender
Household structures 
Marital Status
Demographic Dynamics

# 3.I envisioned that the scope of this research should span from the micro level—the household unit—to the macro level—the demographic and national composition of the population—and ultimately to the dynamics of population mobility across the entire city. However, since my specific task was to formulate one research question, I chose to approach the subject solely from a micro-level perspective.

# 4.But,we noticed that: Our data consists of aggregated data,therefore, we cannot draw causal inferences at the individual level.And our data lacks cross-dimensions and joint variables.

# 5.Try formulating a question.
```{r}
#1. 近年来家庭规模的变化趋势（家庭人数，居住规模-年龄分布，婚姻法律状态）验证老龄化的影响，新价值观念的影响。
#验证已知理论在慕尼黑这个特定地方的具体表现。新结论是两种力量的结构比例和时间拐点。
#Haushalte nach Personenzahl × Altersgruppen × Familienstand。
#1.1（描述性问题）过去 16 年间，慕尼黑各个规模家庭（从 1 人户到 5人以上户）的整体结构比例发生了怎样的迁移？是否存在明显的“家庭规模重心向 1-2 人户倾斜”的趋势？有无特定年份转折点？
#1.2（机制拆解）随着人均寿命的延长和人口红利的消退，慕尼黑独居老人（65岁以上且法律状态为丧偶 verwitwet 或离异 geschieden）的数字呈现怎样的增长？这一群体在整个慕尼黑一人户家庭的增长份额中占据了多大比例？
#1.3（机制拆解）现代新价值观念（如非婚同居）如何体现在慕尼黑的家庭形式中？在过去 16 年中，未婚人口（ledig）的增长是否与两人户（2 Personen）的增长呈现同步趋势，从而验证慕尼黑中青年“伴侣同住但不结婚”的第二次人口转型特征？
```

# 6. First I would like to know whether the results yielded by my research questions can validate my "sociological" theory. If it cannot, maybe I will change a question. So I first let Gemini to draw me a few line pictures from the real dataframe. But I won’t use its code to our project. I will finally write code by myself.
```{r}
# clean the data
library(tidyverse)
library(lubridate)

# 1. 先揪出到底是谁没转换成功，顺便清洗数据
data_cleaned <- data %>%
  # 先把字符强制转成数字
  mutate(
    JAHR = as.numeric(JAHR),
    WERT = as.numeric(WERT),
    MONAT_NUM = as.numeric(MONAT)
  ) %>%
  # 关键修复：过滤掉那些不是正常月份的特殊行（比如末尾是00的年度总计行）
  # 正常的年月应该是 202401 到 202412，所以 MONAT % 100 应该在 1 到 12 之间
  filter((MONAT_NUM %% 100) >= 1 & (MONAT_NUM %% 100) <= 12) %>%
  # 排除掉之后，再放心地转换时间
  mutate(
    Date = ymd(paste0(MONAT, "01"))
  )

# 2. 检查一下是不是还有没转换成功的 NA
cat("检查：现在还有多少个时间是转换失败的？ ", sum(is.na(data_cleaned$Date)), "个\n")
```
```{r}
# Examining Trends in Household Size Over Recent Years
# ==========================================
# 运行 1.1 的画图代码
# ==========================================

# 1. 筛选家庭户数大类
theme1_1 <- data_cleaned %>%
  filter(MONATSZAHL == "Haushalte nach Personenzahl") %>%
  filter(AUSPRAEGUNG %in% c("1 Person", "2 Personen", "3 Personen", "4 Personen", "5 Personen und mehr")) %>%
  filter(month(Date) == 12) # 取每年12月，让长线趋势更平滑

# 2. 绘制绝对数量折线图
ggplot(theme1_1, aes(x = Date, y = WERT, color = AUSPRAEGUNG)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "1.1 慕尼黑不同规模家庭绝对数量发展趋势 (2010-2026)",
       x = "年份", y = "家庭总数 (户)", color = "家庭规模")
```
```{r}
# Examining whether population aging has contributed to the shrinking of family size.
# ==========================================
# 运行 1.2 的画图代码 (英文版防乱码)
# ==========================================

# 1. 筛选一人户与老龄化相关的平行标签
theme1_2 <- data_cleaned %>%
  filter(AUSPRAEGUNG %in% c("1 Person", "Rentner*innen (65 Jahre und älter)", "verwitwet", "geschieden")) %>%
  filter(month(Date) == 12)

# 2. 绘制分盘趋势拟合图
ggplot(theme1_2, aes(x = Date, y = WERT, color = AUSPRAEGUNG)) +
  geom_line(size = 1) +
  geom_point(size = 1.5) +
  facet_wrap(~AUSPRAEGUNG, scales = "free_y", ncol = 2) + 
  theme_minimal(base_size = 11) +
  labs(title = "1.2 Trend Alignment: 1-Person Households vs. Ageing Indicators",
       x = "Year", y = "Absolute Value (Count)") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 10))
```

```{r}
# Examining the Association Between the Non-marital Cohabitation Hypothesis and Family Size
# ==========================================
# 运行 1.3 的画图代码 (英文版防乱码)
# ==========================================

# 1. 筛选非婚同居假说相关的平行标签
theme1_3 <- data_cleaned %>%
  filter(AUSPRAEGUNG %in% c("2 Personen", "ledig", "verheiratet")) %>%
  filter(month(Date) == 12)

# 2. 绘制分盘趋势图
ggplot(theme1_3, aes(x = Date, y = WERT, color = AUSPRAEGUNG)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~AUSPRAEGUNG, scales = "free_y", ncol = 1) + # 纵向排列，最方便对比时间节点
  theme_minimal(base_size = 11) +
  labs(title = "1.3 Nuptial Shift: 2-Person Households vs. Marital Status",
       x = "Year", y = "Absolute Value (Count)") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 10))
```
# 7.I have found that family size has not contracted due to aging as much as I had anticipated, nor has it been significantly influenced by new societal values. In recent years, family size has remained relatively stable; the only exception is single-person households, which have shown a significant upward trend.

# 8.Therefore, I would like to shift my focus in a different direction—specifically, to examine why the prevalence of single-person households is on the rise. To do so, I will turn my attention to the role of immigration.

```{r}
# Examining trends in the number of EU and non-EU nationals
# 1. 提取一人户 (严格限制在 2024 年及以前)
df_1person <- data_cleaned %>%
  filter(AUSPRAEGUNG == "1 Person" & month(Date) == 12 & JAHR <= 2024) %>%
  select(Date, WERT) %>%
  mutate(Label = "1 Person (Single Households)")

# 2. 提取传统德国人
df_native <- data_cleaned %>%
  filter(AUSPRAEGUNG == "deutsch ohne Migrationshintergrund" & month(Date) == 12 & JAHR <= 2024) %>%
  select(Date, WERT) %>%
  mutate(Label = "Native Germans (Without Migration Background)")

# 3. 聚合非欧盟总数 (剔除2025-2026的统计未完成年度)
df_non_eu <- data_cleaned %>%
  filter(MONATSZAHL == "Nicht-EU-Staatsangehörigkeiten" & month(Date) == 12 & JAHR <= 2024) %>%
  group_by(Date) %>%
  summarise(WERT = sum(WERT, na.rm = TRUE)) %>% 
  ungroup() %>%
  mutate(Label = "Total Non-EU Citizens (International Inflows)")

# 4. 合体
exploratory_final <- bind_rows(df_1person, df_native, df_non_eu)

# 5. 绘图
ggplot(exploratory_final, aes(x = Date, y = WERT, color = Label)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  facet_wrap(~Label, scales = "free_y", ncol = 1) + 
  theme_minimal(base_size = 11) +
  labs(title = "Exploratory Verification: The Demographic Substitution Effect in Munich (2000-2024)",
       x = "Year", y = "Absolute Population / Households") +
  theme(legend.position = "none",
        strip.text = element_text(face = "bold", size = 10))
```
# 9.I have observed that the number of non-EU residents in recent years has been on the rise, while the number of local Germans has declined. Concurrently, the prevalence of single-person households has risen in parallel with the increase in the non-EU population. I therefore conclude that an increasing number of migrants—such as international students and skilled professionals—are moving to Munich, and that these newcomers are the primary contributors to the growth in single-person households.

# 10.This outlines the process and logic I followed to define my research problem. Subsequent projects will build upon this foundation, and I will be writing the code myself.
